Measurement Bias from Address Aliasing

### 2014-08-13 paper.tex
Add table describing test machines

\section{Introduction}
Accurately measuring the performance of computer programs is important in order to evaluate new algorithms, or compare different implementations.
A key challenge for performance analysts and systems researchers is handling \emph{measurement bias}.
Changes to external factors of the system, such as link ordering or the contents of environment variables, has been shown to have potentially significant impacts on performance in real applications~\cite{Mytkowicz:2009:WrongData}.
The underlying explanation is likely that this changes memory addresses of code or data, which in turn interacts with various low level hardware buffers at runtime.
Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes to for example system environment variables ultimately affect program execution.

\pgfplotstableread{bin/microkernel-cycles-core2.dat}{\microkernelcoretwotable}

\begin{figure}[t]
\begin{minipage}[t]{0.40\textwidth}
  \vspace{3mm}
  \begin{lstlisting}[frame=tb] %[frame=single]
  static int i, j, k;
  int main() {
      int g = 0, inc = 1;
      for (; g < 65536; g++) {
          i += inc;
          j += inc;
          k += inc;
      }
      return 0;
  }
  \end{lstlisting}
\end{minipage}
%\hfill
\begin{minipage}[t]{0.55\textwidth}
  \begin{flushleft}
    %\fbox{
    \begin{tikzpicture}
      \begin{axis}[
          width = 7.8cm,
          height = 4.5cm,
          font = \footnotesize,
          xlabel=Bytes added to environment,
          ylabel=Cycles,
          domain = 0:4096,
          xtick = {0,1024,...,4096},
          xmin = 0,
          xmax = 4096,
          cycle list name = exotic
        ]
        \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \microkernelcoretwotable ;
      \end{axis}
    \end{tikzpicture}
    %}
  \end{flushleft}
\end{minipage}
\label{fig:motivation}
\caption{Microkernel first presented in~\cite{Mytkowicz:2009:WrongData}, showing bias to environment size}
\end{figure}

@@

Because bias can be hardware dependent, and to keep the scope of this work manageable, we choose to focus on the Intel ``Haswell'' microarchitecture specifically.
Our setup consists of a 4th generation Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-4770K processor, running 64-bit Ubuntu 14.04 LTS on kernel version 3.13.0-24-generic.
Code samples are compiled using the {\small GCC} toolchain, version 4.8.2-19ubuntu1.

@@

While optimizations such as these are good on the average case, there are corner cases. 
In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096 bytes.
A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
The number of times this happens can be counted by the following performance counter:
\begin{description}
  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}] ``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.'' Intel Optimization Manual~\citeyear[B.3.4.4]{OptimizationManual}
\end{description}

>>>>

\section{Introduction}
Accurately measuring the performance of computer programs is important in order to evaluate new algorithms, or compare different implementations.
A key challenge for performance analysts and systems researchers is handling \emph{measurement bias}.
Changes to external factors of the system, such as link ordering or the contents of environment variables, has been shown to have potentially significant impacts on performance in real applications~\cite{Mytkowicz:2009:WrongData}.
The underlying explanation is likely that this changes memory addresses of code or data, which in turn interacts with various low level hardware buffers at runtime.
Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes to for example system environment variables ultimately affect program execution.
In Figure~\ref{fig:motivation} we have reproduced a result originally presented by \citeN{Mytkowicz:2009:WrongData}, showing how the number of cycles executed for a simple program depends on environment variables.
This example provides the basis for our analysis.

\pgfplotstableread{bin/microkernel-cycles-core2.dat}{\microkernelcoretwotable}
\begin{figure}[t]
\begin{minipage}[t]{0.40\textwidth}
  \vspace{3mm}
  \begin{lstlisting}[frame=tb] %[frame=single]
  static int i, j, k;
  int main() {
      int g = 0, inc = 1;
      for (; g < 65536; g++) {
          i += inc;
          j += inc;
          k += inc;
      }
      return 0;
  }
  \end{lstlisting}
\end{minipage}
%\hfill
\begin{minipage}[t]{0.55\textwidth}
  \begin{flushleft}
    %\fbox{
    \begin{tikzpicture}
      \begin{axis}[
          title = Core2 Duo,
          width = 7.8cm,
          height = 4.5cm,
          font = \footnotesize,
          xlabel=Bytes added to environment,
          ylabel=Cycles,
          domain = 0:4096,
          xtick = {0,1024,...,4096},
          xmin = 0,
          xmax = 4096,
          cycle list name = exotic
        ]
        \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \microkernelcoretwotable ;
      \end{axis}
    \end{tikzpicture}
    %}
  \end{flushleft}
\end{minipage}
\label{fig:motivation}
\caption{Microkernel first presented by \citeN{Mytkowicz:2009:WrongData}, showing bias to environment size}
\end{figure}

@@

Because bias can be hardware dependent, and to keep the scope of this work manageable, we initially chose to focus our analysis on the 4th generation Intel ``Haswell'' microarchitecture specifically.
We later reproduced similar results on the previous ``Sandy Bridge'' CPU, as well as an older Core2 Duo machine.
Each setup is described in Table~\ref{tab:machines}, and referenced throughout the paper.

\begin{table}
  \tbl{Experimental setups used\label{tab:machines}}{
  \begin{tabular}{|l|l|}
    \hline
Core2 Duo     & Intel\textsuperscript{\textregistered{}} Core\texttrademark{}2 Duo P8600 (laptop) \\
              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0), GCC 4.8.2-19ubuntu1 \\ \hline
Sandy Bridge  & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-3???? \\
              & Ubuntu ?? \\ \hline
Haswell       & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-4770K \\
              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0), GCC 4.8.2-19ubuntu1 \\ \hline
  \end{tabular}}
\end{table}

@@

While optimizations such as these are good on the average case, there are corner cases. 
In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096 bytes.
A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
On recent architectures such as Haswell and Ivy Bridge, the number of times this happens can be counted by the following performance counter:
\begin{description}
  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}] ``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.'' Intel Optimization Manual~\citeyear[B.3.4.4]{OptimizationManual}
\end{description}
Older architectures such as Core2 does not have the same event available.
However, the effect is still described in the manual for these processors, and covered by the following event:
\begin{description}
  \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE}] ``(...) Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.'' ~\citeN[Table 19-17]{Volume3B}
\end{description}
